---
title: roy_deblina  
output:
  html_document: default
---

```{r setup, include=FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

library(moments)  # install.packages("moments")

```
-----

### Test Items starts from here - There are 5 sections - 75 points total ##########################

##### Section 1: (15 points) ##################################

##### (1) R has probability functions available for use (Kabacoff, Section 5.2.3). Using one distribution to approximate another is not uncommon.

(1)(a) (6 points) The Poisson distribution may be used to approximate the binomial distribution if n > 20 and np < 7. Estimate the following binomial probabilities using *dpois()* or *ppois()* with probability p = 0.05, and n = 100. Then, estimate the same probabilities using *dbinom()* or *pbinom()*.  Show the numerical results of your calculations.

(i) The probability of exactly 0 successes.

```{r test1ai}
# Step 1: Define the number of trials and probability of success
n <- 100         # number of trials
p <- 0.05        # probability of success
lambda <- n * p  # Poisson mean (λ = n * p)

# Step 2: Calculate P(X = 0) using Poisson approximation
poisson_p0 <- dpois(0, lambda)  # P(X = 0) using Poisson

# Step 3: Calculate P(X = 0) using exact Binomial distribution
binomial_p0 <- dbinom(0, n, p)  # P(X = 0) using Binomial

# Step 4: Print the results, rounded to 5 decimal places
cat("Poisson Approximation (P(X = 0)):", round(poisson_p0, 5), "\n")
cat("Exact Binomial (P(X = 0))       :", round(binomial_p0, 5), "\n")


```

(ii) The probability of fewer than 6 successes. Please note the following, taken from the Binomial Distribution R Documentation page, regarding the "lower.tail" argument:

lower.tail	logical; if TRUE (default), probabilities are P[X ??? x], otherwise, P[X > x].

```{r test1aii}
# Parameters
n <- 100         # number of trials
p <- 0.05        # success probability
lambda <- n * p  # Poisson mean

# (ii) Probability of fewer than 6 successes (i.e., P(X < 6) = P(X ≤ 5))

# Poisson approximation
poisson_p_less6 <- ppois(5, lambda, lower.tail = TRUE)

# Exact binomial probability
binomial_p_less6 <- pbinom(5, n, p, lower.tail = TRUE)

# Display results
cat("Poisson Approximation (P(X < 6)):", round(poisson_p_less6, 5), "\n")
cat("Exact Binomial (P(X < 6))       :", round(binomial_p_less6, 5), "\n")

```

The binomial may also be approximated via the normal distribution. Estimate the following binomial probabilities using *dnorm()* or *pnorm()*, this time with probability p = 0.25 and n = 100. Then, calculate the same probabilities using *dbinom()* and *pbinom()*. Use continuity correction. Show the numerical results of your calculations.

(iii) The probability of exactly 25 successes.

```{r test1aiii}
# Parameters
n <- 100
p <- 0.25
mean <- n * p
sd <- sqrt(n * p * (1 - p))

# (iii) Probability of exactly 25 successes

# Binomial exact
binom_exact_25 <- dbinom(25, size = n, prob = p)

# Normal approximation with continuity correction: P(24.5 < X < 25.5)
normal_approx_25 <- pnorm(25.5, mean, sd) - pnorm(24.5, mean, sd)

# Display the results
cat("Normal Approximation (P(X = 25)) :", round(normal_approx_25, 5), "\n")
cat("Exact Binomial (P(X = 25))       :", round(binom_exact_25, 5), "\n")

```

(iv) The probability of fewer than 20 successes. Please note the following, taken from the Normal Distribution R Documentation page, regarding the "lower.tail" argument:

lower.tail	logical; if TRUE (default), probabilities are P[X ??? x], otherwise, P[X > x].

```{r test1aiv}
# Parameters
n <- 100
p <- 0.25
mean <- n * p
sd <- sqrt(n * p * (1 - p))

# (iv) Probability of fewer than 20 successes

# Binomial exact: P(X ≤ 19)
binom_less20 <- pbinom(19, size = n, prob = p, lower.tail = TRUE)

# Normal approximation with continuity correction: P(X ≤ 19.5)
normal_less20 <- pnorm(19.5, mean = mean, sd = sd, lower.tail = TRUE)

# Display results
cat("Normal Approximation (P(X < 20)) :", round(normal_less20, 5), "\n")
cat("Exact Binomial (P(X < 20))       :", round(binom_less20, 5), "\n")

```

(1)(b) (3 points) Generate side-by-side barplots using *par(mfrow = c(1,2))* or *grid.arrange()*. The left barplot will show Poisson probabilties for outcomes ranging from 0 to 10. The right barplot will show binomial probabilities for outcomes ranging from 0 to 10. Use p = 0.05 and n = 100. Title each plot,  present in color and assign names to the bar; i.e. x-axis value labels.

```{r test1b}
# Set up the layout: 1 row, 2 columns
par(mfrow = c(1, 2))

# Define parameters
n <- 100
p <- 0.05
lambda <- n * p
x_vals <- 0:10

# Calculate probabilities
pois_probs <- dpois(x_vals, lambda)
binom_probs <- dbinom(x_vals, size = n, prob = p)

# Poisson Barplot (Left)
barplot(pois_probs,
        names.arg = x_vals,
        col = "skyblue",
        main = "Poisson Probabilities\n(λ = 5)",
        xlab = "Number of Successes",
        ylab = "Probability",
        ylim = c(0, max(c(pois_probs, binom_probs)) + 0.02))

# Binomial Barplot (Right)
barplot(binom_probs,
        names.arg = x_vals,
        col = "orange",
        main = "Binomial Probabilities\n(n = 100, p = 0.05)",
        xlab = "Number of Successes",
        ylab = "Probability",
        ylim = c(0, max(c(pois_probs, binom_probs)) + 0.02))

```

(1)(c) (6 points): For this problem, refer to Sections 5.2 of Business Statistics. A discrete random variable has outcomes:  0, 1, 2, 3, 4, 5, 6.  The corresponding probabilities in sequence with the outcomes are: 0.214, 0.230, 0.240, 0.182, 0.130, 0.003, 0.001.  In other words, the probabilty of obtaining "0" is 0.215.  

(i) Calculate the expected value and variance for this distribution using the general formula for mean and variance of a discrete distribution. To do this, you will need to use integer values from 0 to 6 as outcomes along with the corresponding probabilities. Round your answer to 2 decimal places. 

```{r test1ci}
# Define outcomes and probabilities
x <- 0:6
p <- c(0.214, 0.230, 0.240, 0.182, 0.130, 0.003, 0.001)

# Calculate expected value (mean)
expected_value <- sum(x * p)

# Calculate variance
variance <- sum((x^2) * p) - expected_value^2

# Round results to 2 decimal places
expected_value <- round(expected_value, 2)
variance <- round(variance, 2)

# Display results
cat("Expected Value (Mean):", expected_value, "\n")
cat("Variance:", variance, "\n")

```

(ii) Use the *cumsum()* function and plot the cumulative probabilties versus the corresponding outcomes. Detemine the value of the median for this distribution and show on this plot. Note that there are methods for interpolating a median. However, we can identify an appropriate median from our set of our outcomes - 0 through 6 - that satisfies the definition. Creating a stair-step plot of the cumulative probability as a function of the outcomes may be helpful in identifying it.

```{r test1cii}
# Define outcomes and probabilities
x <- 0:6
p <- c(0.214, 0.230, 0.240, 0.182, 0.130, 0.003, 0.001)

# Compute cumulative probabilities
cum_probs <- cumsum(p)

# Find the median: first x for which cumulative probability ≥ 0.5
median_x <- x[which(cum_probs >= 0.5)[1]]

# Plot the cumulative probabilities as a stair-step function
plot(x, cum_probs, type = "s", col = "blue", lwd = 2,
     xlab = "Outcome", ylab = "Cumulative Probability",
     main = "Cumulative Probability Distribution")

# Add a horizontal reference line at 0.5
abline(h = 0.5, col = "darkgray", lty = 2)

# Add a vertical line at the median
abline(v = median_x, col = "red", lty = 2)

# Annotate the median
text(median_x + 0.2, 0.55, paste("Median =", median_x), col = "red")


```

##### Section 2: (15 points) ##################################

#####  (2) Conditional probabilities appear in many contexts and, in particular, are used by Bayes' Theorem. Correlations are another means for evaluating dependency between variables. The dataset "faithful"" is part of the "datasets" package and may be loaded with the statement *data(faithful)*. It contains 272 observations of 2 variables;  waiting time between eruptions (in minutes) and the duration of the eruption (in minutes) for the Old Faithful geyser in Yellowstone National Park.

(2)(a) (6 points) Load the "faithful" dataset and present summary statistics and a histogram of waiting times.  Additionally, compute the empirical conditional probability of an eruption less than 3.0 minutes, if the waiting time exceeds 70 minutes.

```{r test2a}
data(faithful, package = "datasets")
# 1. Summary statistics for 'waiting' times
summary_stats <- summary(faithful$waiting)
print(summary_stats)

# 2. Histogram of waiting times
hist(faithful$waiting,
     col = "lightblue",
     main = "Histogram of Waiting Times",
     xlab = "Waiting time (minutes)",
     border = "black")

# 3. Compute empirical conditional probability:
#    P(duration < 3.0 | waiting > 70)

# Subset the data for waiting > 70
subset_data <- faithful[faithful$waiting > 70, ]

# Count number of eruptions < 3.0 minutes in that subset
num_eruption_lt3 <- sum(subset_data$eruptions < 3.0)

# Total number of eruptions where waiting > 70
total_wait_gt70 <- nrow(subset_data)

# Empirical conditional probability
cond_prob <- num_eruption_lt3 / total_wait_gt70

# Print result rounded to 3 decimals
cat("Empirical P(eruption < 3.0 | waiting > 70):", round(cond_prob, 3), "\n")


```

(i) Identify any observations in "faithful" for which the waiting time exceeds 70 minutes and the eruptions are less than 3.0 minutes.  List and show any such observations in a distinct color on a scatterplot of all eruption (vertical axis) and waiting times (horizontal axis). Include a horizontal line at eruption = 3.0, and a vertical line at waiting time = 70.  Add a title and appropriate text. 

```{r test2ai}
 # Load the dataset
data(faithful)

# Step 1: Identify observations where waiting > 70 and eruption < 3.0
subset_outliers <- faithful[faithful$waiting > 70 & faithful$eruptions < 3.0, ]

# Step 2: Display these observations
print(subset_outliers)

# Step 3: Create scatterplot
plot(faithful$waiting, faithful$eruptions,
     main = "Eruption Duration vs Waiting Time\n(Old Faithful Geyser)",
     xlab = "Waiting Time (minutes)",
     ylab = "Eruption Duration (minutes)",
     col = "grey", pch = 19)

# Step 4: Add horizontal and vertical reference lines
abline(h = 3.0, col = "darkblue", lty = 2)   # eruption = 3.0
abline(v = 70, col = "darkblue", lty = 2)    # waiting = 70

# Step 5: Highlight identified outliers in red
points(subset_outliers$waiting, subset_outliers$eruptions,
       col = "red", pch = 19, cex = 1.2)

# Step 6: Add a label
text(80, 2.5, "Eruption < 3.0 & Waiting > 70", col = "red", pos = 4)
  
```

(ii) What does the plot suggest about the relationship between eruption time and waiting time?

***Answer: 
The plot suggests a positive relationship between eruption duration and waiting time. Specifically:
Longer waiting times tend to lead to longer eruptions.
Shorter waiting times are generally associated with shorter eruptions.
The data appears to form two clusters, indicating a bimodal pattern - possibly representing two types of geyser behavior.
This implies that the geyser may operate in two distinct cycles, and eruption duration is not random, but rather depends on the waiting time before it.

In summary: As waiting time increases, eruption duration also increases.***


(2)(b) (6 points) Past research indicates that the waiting times between consecutive eruptions are not independent.  This problem will check to see if there is evidence of this. Form consecutive pairs of waiting times.  In other words, pair the first and second waiting times, pair the third and fourth waiting times, and so forth.  There are 136 resulting consecutive pairs of waiting times.  Form a data frame with the first column containing the first waiting time in a pair and the second column with the second waiting time in a pair. Plot the pairs with the second member of a pair on the vertical axis and the first member on the horizontal axis.

One way to do this is to pass the vector of waiting times - faithful$waiting - to *matrix()*, specifying 2 columns for our matrix, with values organized by row; i.e. byrow = TRUE.

```{r test2b}
# Step 1: Load the dataset
data(faithful)

# Step 2: Extract the 'waiting' time vector
waiting_times <- faithful$waiting

# Step 3: Form consecutive pairs using matrix (by rows)
#         matrix() reshapes the vector into 2 columns and fills by row
pair_matrix <- matrix(waiting_times, ncol = 2, byrow = TRUE)

# Step 4: Convert the matrix into a data frame with clear column names
wait_pairs <- data.frame(
  Wait1 = pair_matrix[, 1],  # first waiting time in the pair
  Wait2 = pair_matrix[, 2]   # second waiting time in the pair
)

# Step 5: Plot the pairs (Wait1 on x-axis, Wait2 on y-axis)
plot(wait_pairs$Wait1, wait_pairs$Wait2,
     main = "Consecutive Waiting Time Pairs",
     xlab = "First Waiting Time (minutes)",
     ylab = "Second Waiting Time (minutes)",
     col = "darkgreen", pch = 19)
```

(2)(c) (3 points) Test the hypothesis of independence with a two-sided test at the 5% level using the Kendall correlation coefficient. The *cor.test()* function can be used to structure this test and specify the appropriate - Kendall's tau - method.

```{r test2c}
# Step 1: Load the dataset and form pairs (if not done already)
data(faithful)
waiting_times <- faithful$waiting
pair_matrix <- matrix(waiting_times, ncol = 2, byrow = TRUE)
wait_pairs <- data.frame(
  Wait1 = pair_matrix[, 1],
  Wait2 = pair_matrix[, 2]
)

# Step 2: Perform Kendall's tau correlation test
kendall_test <- cor.test(wait_pairs$Wait1, wait_pairs$Wait2, method = "kendall")

# Step 3: Print the test result
print(kendall_test)

```

##### Section 3: (15 points) ##################################

##### (3)  Performing hypothesis tests using random samples is fundamental to statistical inference. The first part of this problem involves comparing two different diets. Using "ChickWeight" data available in the base R, "datasets" package, we will create a subset of the "ChickWeight" data frame. Specifically, we want to create a data frame that includes only those rows where Time == 21 AND Diet == 1 or 3.

```{r test3}

# load "ChickWeight" dataset
data(ChickWeight, package = "datasets")

# There are multiple ways to approach the subsetting task. The method you choose is up
# to you.

# Subset the data for Time = 21 and Diet = 1 or 3
df <- subset(ChickWeight, Time == 21 & (Diet == 1 | Diet == 3))

# View the first few rows to confirm the match
head(df)

# The values in your subsetted data frame should match those below:
# > head(df)
#    weight Time Chick Diet
# 12    205   21     1    1
# 24    215   21     2    1
# 36    202   21     3    1
# 48    157   21     4    1
# 60    223   21     5    1
# 72    157   21     6    1

```

##### The data frame, "result", has chick weights for two diets, identified as diet "1" and "3". Use the data frame, "result," to complete the following item.

(3)(a) (3 points) Display two side-by-side vertical boxplots using par(mfrow = c(1,2)).  One boxplot would display Diet "1" and the other Diet "3". 

```{r test3a}
# Set plotting area to 1 row, 2 columns
par(mfrow = c(1, 2))

# Boxplot for Diet 1
boxplot(weight ~ Diet, data = df[df$Diet == 1, ],
        main = "Diet 1",
        ylab = "Weight",
        col = "lightblue",
        border = "darkblue")

# Boxplot for Diet 3
boxplot(weight ~ Diet, data = df[df$Diet == 3, ],
        main = "Diet 3",
        ylab = "Weight",
        col = "lightgreen",
        border = "darkgreen")

```

(3)(b) (3 points)  Use the "weight" data for the two diets to test the null hypothesis of equal population mean weights for the two diets. Test at the 95% confidence level with a two-sided t-test. This can be done using *t.test()* in R. Assume equal variances. Display the results of *t.test()*.

```{r test3b}

# Create vectors for weights of Diet 1 and Diet 3
diet1_weights <- df$weight[df$Diet == 1]
diet3_weights <- df$weight[df$Diet == 3]

# Perform two-sample t-test (equal variances assumed)
ttest_result <- t.test(diet1_weights, diet3_weights,
                       var.equal = TRUE,        # assume equal variances
                       alternative = "two.sided",  # two-sided test
                       conf.level = 0.95)       # 95% confidence

# Step 4: Display the result
print(ttest_result)

```

##### Working with paired data is another common statistical activity. The "ChickWeight" data will be used to illustrate how the weight gain from day 20 to 21 may be analyzed. This time, we will look only at those individuals on Diet == "3". You will need to add code below creating two (2) vectors. One (1) vector should include all the Time == 20 weights of those individuals on Diet == "3"; a second should include all the Time == 21 weights of those individuals on Diet == "3".

```{r test3paired}

# There are multiple ways to approach the subsetting task. The method you choose is up
# to you.
# Step 1: Load the dataset
data(ChickWeight)

# Step 2: Filter data for Diet == 3 and Time in 20 or 21
diet3_data <- ChickWeight[ChickWeight$Diet == 3 & ChickWeight$Time %in% c(20, 21), ]

# Step 3: Split into Day 20 and Day 21 data
day20 <- diet3_data[diet3_data$Time == 20, c("Chick", "weight")]
day21 <- diet3_data[diet3_data$Time == 21, c("Chick", "weight")]

# Step 4: Merge by Chick ID to ensure pairing
paired <- merge(day20, day21, by = "Chick", suffixes = c("_day20", "_day21"))

# Step 5: Sort by Chick ID to match expected order
paired <- paired[order(as.numeric(as.character(paired$Chick))), ]

# Step 6: Extract vectors
weight_day20 <- paired$weight_day20
weight_day21 <- paired$weight_day21

# Optional: View first 6 weights on Day 20
head(weight_day20)

# The first six (6) elements of your Time == 20 vector should match those below:
# [1] 235 291 156 327 361 225

```

(3)(c) (3 points) Present a scatterplot of the Time == 21 weights as a function of the Time == 20 weights.  Include a diagonal line with zero intercept and slope equal to one. Title and label the variables in this scatterplot.  

```{r test3c}
# Create the scatterplot
plot(weight_day20, weight_day21,
     main = "Chick Weights: Day 21 vs Day 20 (Diet 3)",
     xlab = "Weight on Day 20 (grams)",
     ylab = "Weight on Day 21 (grams)",
     pch = 19, col = "darkblue")

# Add reference diagonal line y = x
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)


```

(3)(d) (6 points) Calculate and present a one-sided, 95% confidence interval for the average weight gain from day 20 to day 21. Write the code for the paired t-test and for determination of the confidence interval endpoints. **Do not use *t.test()**, although you may check your answers using this function. Present the resulting test statistic value, critical value, p-value and confidence interval.

```{r test3d}
# Calculate weight gains
gain <- weight_day21 - weight_day20

# Sample size
n <- length(gain)

# Sample mean and standard deviation of the differences
mean_gain <- mean(gain)
sd_gain <- sd(gain)

# Test statistic
t_stat <- mean_gain / (sd_gain / sqrt(n))

# Critical t-value for one-sided test at 95% confidence (alpha = 0.05)
t_crit <- qt(0.95, df = n - 1)

# p-value for one-sided test
p_value <- 1 - pt(t_stat, df = n - 1)

# Lower bound of one-sided 95% confidence interval
ci_lower <- mean_gain - t_crit * (sd_gain / sqrt(n))

# Output the results
cat("Sample size (n):", n, "\n")
cat("Mean weight gain:", round(mean_gain, 2), "\n")
cat("Standard deviation:", round(sd_gain, 2), "\n")
cat("Test statistic (t):", round(t_stat, 3), "\n")
cat("Critical value (t_crit @ 95%):", round(t_crit, 3), "\n")
cat("p-value (one-sided):", round(p_value, 4), "\n")
cat("One-sided 95% Confidence Interval: [", round(ci_lower, 2), ", Inf )\n")


```

##### Section 4: (15 points) ##################################

##### (4) Statistical inference depends on using a sampling distribution for a statistic in order to make confidence statements about unknown population parameters. The Central Limit Theorem is used to justify use of the normal distribution as a sampling distribution for statistical inference. Using Nile River flow data from 1871 to 1970, this problem demonstrates sampling distribution convergence to normality. Use the code below to prepare the data.  Refer to this example when completing (4)(c) below.

```{r test4}
data(Nile, package = "datasets")


```

(4)(a) (3 points) Using Nile River flow data and the "moments" package, calculate skewness and kurtosis. Present a QQ plot and boxplot of the flow data side-by-side using *qqnorm()*, *qqline()* and *boxplot()*; *par(mfrow = c(1, 2))* may be used to locate the plots side-by-side. Add features to these displays as you choose.

```{r test4a}
# Step 1: Load required packages and data
library(moments)
data(Nile, package = "datasets")

# Step 2: Convert to numeric vector (optional but safe)
nile_data <- as.numeric(Nile)

# Step 3: Calculate skewness and kurtosis
skew_val <- skewness(nile_data)
kurt_val <- kurtosis(nile_data)

# Step 4: Display the skewness and kurtosis
cat("Skewness of Nile River flow:", round(skew_val, 3), "\n")
cat("Kurtosis of Nile River flow:", round(kurt_val, 3), "\n")

# Step 5: Plot QQ plot and boxplot side-by-side
par(mfrow = c(1, 2))  # Set 1 row, 2 columns

# QQ plot
qqnorm(nile_data,
       main = "QQ Plot: Nile River Flow",
       col = "blue",
       pch = 19)
qqline(nile_data, col = "red", lwd = 2)
legend("topleft", legend = "QQ Line", col = "red", lty = 1, bty = "n")

# Boxplot
boxplot(nile_data,
        main = "Boxplot: Nile River Flow",
        ylab = "Flow (10^8 m³)",
        col = "lightgreen",
        border = "darkgreen",
        notch = TRUE)


```

(4)(b) (6 points) Using *set.seed(124)* and the Nile data, generate 1000 random samples of size n = 16, with replacement. For each sample drawn, calculate and store the sample mean. This can be done with a for-loop and use of the *sample()* function. Label the resulting 1000 mean values as "sample1". **Repeat these steps using *set.seed(127)* - a different "seed" - and samples of size n = 64.** Label these 1000 mean values as "sample2". Compute and present the means, sample standard deviations and sample variances for "sample1" and "sample2" in a table with the first row for "sample1", the second row for "sample2" and the columns labled for each statistic.

```{r test4b}
# Load the data again
data(Nile, package = "datasets")
nile_data <- as.numeric(Nile)

# Initialize sample1 with set.seed(124) and sample size 16
set.seed(124)
sample1 <- numeric(1000)

for (i in 1:1000) {
  sample1[i] <- mean(sample(nile_data, size = 16, replace = TRUE))
}

# Initialize sample2 with set.seed(127) and sample size 64
set.seed(127)
sample2 <- numeric(1000)

for (i in 1:1000) {
  sample2[i] <- mean(sample(nile_data, size = 64, replace = TRUE))
}

# Compute summary statistics
mean1 <- mean(sample1)
sd1 <- sd(sample1)
var1 <- var(sample1)

mean2 <- mean(sample2)
sd2 <- sd(sample2)
var2 <- var(sample2)

# Create summary table
result_table <- data.frame(
  Sample = c("sample1", "sample2"),
  Mean = round(c(mean1, mean2), 2),
  Standard_Deviation = round(c(sd1, sd2), 2),
  Variance = round(c(var1, var2), 2)
)

# Print the table
print(result_table, row.names = FALSE)


```

(4)(c) (6 points) Present side-by-side histograms of "sample1" and "sample2" with the normal density curve superimposed. To prepare comparable histograms, it will be necessary to use "freq = FALSE" and to maintain the same x-axis with "xlim = c(750, 1050)", and the same y-axis with "ylim = c(0, 0.025)." **To superimpose separate density functions, you will need to use the mean and standard deviation for each "sample" - each histogram - separately.** 

```{r test4c}

# Create histograms of "sample1" and "sample2" with normal density curves superimposed

# Recreate sample1 and sample2 if not already in environment
data(Nile, package = "datasets")
nile_data <- as.numeric(Nile)

set.seed(124)
sample1 <- replicate(1000, mean(sample(nile_data, size = 16, replace = TRUE)))

set.seed(127)
sample2 <- replicate(1000, mean(sample(nile_data, size = 64, replace = TRUE)))

# Compute mean and standard deviation for both samples
mean1 <- mean(sample1)
sd1 <- sd(sample1)

mean2 <- mean(sample2)
sd2 <- sd(sample2)

# Plot side-by-side histograms with normal density curves
par(mfrow = c(1, 2))

# Histogram for sample1
hist(sample1,
     freq = FALSE,
     col = "lightblue",
     main = "Sample Means (n = 16)",
     xlab = "Sample Mean",
     xlim = c(750, 1050),
     ylim = c(0, 0.025),
     border = "white")

curve(dnorm(x, mean = mean1, sd = sd1),
      col = "blue", lwd = 2, add = TRUE)

# Histogram for sample2
hist(sample2,
     freq = FALSE,
     col = "lightgreen",
     main = "Sample Means (n = 64)",
     xlab = "Sample Mean",
     xlim = c(750, 1050),
     ylim = c(0, 0.025),
     border = "white")

curve(dnorm(x, mean = mean2, sd = sd2),
      col = "darkgreen", lwd = 2, add = TRUE)


```

-----

##### Section 5: (15 points) ##################################

##### (5)  This problem deals with contingency table analysis. This is an example of categorical data analysis (see Kabacoff, pp. 145-151). The "warpbreaks" dataset gives the number of warp breaks per loom, where a loom corresponds to a fixed length of yarn.  There are 54 observations on 3 variables: breaks	(numeric, the number of breaks), wool (factor, type of wool: A or B), and tension (factor, low L, medium M and high H).  These data have been studied and used for example elsewhere.  For the purposes of this problem, we will focus on the relationship between breaks and tension using contingency table analysis.

(5)(a)(5 points) warpbreaks is part of the "datasets" package and may be loaded via *data(warpbreaks)*.  Load "warpbreaks" and present the structure using *str()*. Calculate the median number of breaks for the entire dataset, disregarding "tension" and "wool". Define this median value as "median_breaks". Present a histogram of the number of breaks with the location of the median indicated.

Create a new variable "number" as follows:  for each value of "breaks", classify the number of breaks as either strictly below "median_breaks", or the alternative. Convert the "above"|"below" classifications to a factor, and combine with the dataset warpbreaks.  Present a summary of the augmented dataset using *summary()*.  Present a contingency table of the frequency of breaks using the two variables "tension" and "number".  There should be six cells in this table.

```{r test5a}
data(warpbreaks, package = "datasets")

# Display the structure
str(warpbreaks)

# Calculate the median number of breaks (ignoring wool and tension)
median_breaks <- median(warpbreaks$breaks)
cat("Median number of breaks:", median_breaks, "\n")

# Plot histogram of breaks with median line
hist(warpbreaks$breaks,
     main = "Histogram of Warp Breaks",
     xlab = "Number of Breaks",
     col = "lightblue",
     border = "white")
abline(v = median_breaks, col = "red", lwd = 2, lty = 2)
legend("topright", legend = paste("Median =", median_breaks),
       col = "red", lty = 2, bty = "n")

# Create new variable: classify breaks as "below" or "above" median
warpbreaks$number <- ifelse(warpbreaks$breaks < median_breaks, "below", "above")
warpbreaks$number <- factor(warpbreaks$number, levels = c("below", "above"))

# Present summary of augmented dataset
summary(warpbreaks)

# Contingency table of tension vs. number
table(warpbreaks$tension, warpbreaks$number)

```

(5)(b)(3 points)  Using the table constructed in (5)(a), test at the 5% level the null hypothesis of independence using the uncorrected *chisq.test()* (Black, Business Statistics, Section 16.2). Show the results of this test and state your conclusions. 

```{r test5b}
# Construct the contingency table again (if not already in memory)
break_table <- table(warpbreaks$tension, warpbreaks$number)

# Perform Chi-square test of independence (without Yates correction)
chi_test <- chisq.test(break_table, correct = FALSE)

# Display the test result
print(chi_test)


```


(5)(c) (3 points) 'Manually' calculate the chi-squared statistic and p-value of the table from (5)(a). The *addmargins()* function can be used to add row and column sums to the table; useful for calculating the expected values for each cell. You should be able to match the chi-squared and p-values from (5)(b). The underlying code for the *chisq.test()* function can be viewed by entering *chisq.test* - without parentheses - in the Console. You are given code below to create the table, add row and column sums and calculate the expected values for the for the first two (2) of three (3) rows. You will need to add code to calculate the expected values for the third row and the chi-squared. The *pchisq()* function can be used to return the p-value.

```{r test5c}

tbl <- table(warpbreaks$tension, warpbreaks$number)
mar_tbl <- addmargins(tbl)

# For row 1 (Low)
e11 <- mar_tbl[4, 1] * mar_tbl[1, 3] / mar_tbl[4, 3]  # expected for [1,1]
e12 <- mar_tbl[4, 2] * mar_tbl[1, 3] / mar_tbl[4, 3]  # expected for [1,2]

# For row 2 (Medium)
e21 <- mar_tbl[4, 1] * mar_tbl[2, 3] / mar_tbl[4, 3]
e22 <- mar_tbl[4, 2] * mar_tbl[2, 3] / mar_tbl[4, 3]

# For row 3 (High)
e31 <- mar_tbl[4, 1] * mar_tbl[3, 3] / mar_tbl[4, 3]
e32 <- mar_tbl[4, 2] * mar_tbl[3, 3] / mar_tbl[4, 3]

# Step 4: Observed values from tbl
o11 <- tbl[1,1]; o12 <- tbl[1,2]
o21 <- tbl[2,1]; o22 <- tbl[2,2]
o31 <- tbl[3,1]; o32 <- tbl[3,2]

# Compute chi-square statistic manually
chi_sq <- (o11 - e11)^2 / e11 +
          (o12 - e12)^2 / e12 +
          (o21 - e21)^2 / e21 +
          (o22 - e22)^2 / e22 +
          (o31 - e31)^2 / e31 +
          (o32 - e32)^2 / e32

# Compute p-value using pchisq
# Degrees of freedom: (rows - 1) * (cols - 1) = (3 - 1)*(2 - 1) = 2
p_val <- 1 - pchisq(chi_sq, df = 2)

# Output results
cat("Chi-squared statistic:", round(chi_sq, 3), "\n")
cat("Degrees of freedom:", 2, "\n")
cat("p-value:", round(p_val, 4), "\n")

```

(5)(d) (4 points) Build a user-defined function, using your code for (5)(c).We want to pass our (5)(a) table to our function and have it return the chi-squared statistic and p-value. You're provided with the 'shell' of a function and will need to add code to calculate the expected values, the chi-squared statistic, the p-value and return (i.e. output) the chi-squared and p-value.

```{r 5d}

chisq_function <- function(x) {
  # Code for calculating the expected values
  mar_tbl <- addmargins(x)
  e11 <- mar_tbl[4, 1] * mar_tbl[1, 3] / mar_tbl[4, 3]
  e12 <- mar_tbl[4, 2] * mar_tbl[1, 3] / mar_tbl[4, 3]
  e21 <- mar_tbl[4, 1] * mar_tbl[2, 3] / mar_tbl[4, 3]
  e22 <- mar_tbl[4, 2] * mar_tbl[2, 3] / mar_tbl[4, 3]
  e31 <- mar_tbl[4, 1] * mar_tbl[3, 3] / mar_tbl[4, 3]
  e32 <- mar_tbl[4, 2] * mar_tbl[3, 3] / mar_tbl[4, 3]
  
  # Observed values
  o11 <- x[1, 1]; o12 <- x[1, 2]
  o21 <- x[2, 1]; o22 <- x[2, 2]
  o31 <- x[3, 1]; o32 <- x[3, 2]
  
  # Code for calculating the chi-squared
 chi_sq <- (o11 - e11)^2 / e11 +
            (o12 - e12)^2 / e12 +
            (o21 - e21)^2 / e21 +
            (o22 - e22)^2 / e22 +
            (o31 - e31)^2 / e31 +
            (o32 - e32)^2 / e32
  
  # Code for calculating the degrees of freedom and p-value
  # Degrees of freedom = (rows - 1) * (cols - 1)
  df <- (nrow(x) - 1) * (ncol(x) - 1) 
  
   # p-value using the chi-squared distribution
  p_val <- 1 - pchisq(chi_sq, df = df)
    
  # Code to ouput the chi-squared, degrees of freedom and p-value 
   # Return the result as a named list
  return(list(
    Chi_Squared_Statistic = round(chi_sq, 3),
    Degrees_of_Freedom = df,
    P_Value = round(p_val, 4)
  ))
}

chisq_function(tbl)

```

You do not need to do anything with the below. It is provided only for demonstration purposes. In (5)(d), we know the size of the table - 3 x 2 - and write a function to match. Often, though, we'll want to write functions that are flexible in some way.

```{r chisq_vectorized}
# Below is a function that should return the same values as chisq.test() and your
# function from (5)(d). Here, though, the function loops over the rows and columns
# to calculate the expected values. Ideally, this function would work for any sized
# table.

chisqfun <- function(t) {
   x <- addmargins(t)
   e <- matrix(0, nrow = nrow(t), ncol = ncol(t), byrow = T)
   r <- matrix(0, nrow = nrow(t), ncol = ncol(t), byrow = T)
   for (i in 1:dim(t)[1]) {
       for (j in 1:dim(t)[2]) {
          e[i,j] = x[nrow(x),j] * x[i,ncol(x)]/x[nrow(x), ncol(x)]
         r[i,j] = ((x[i,j] - e[i,j])^2)/e[i,j]
         }
     }
  chi <- sum(r)
  xdf <- (nrow(t) - 1) * (ncol(t) - 1)
  pv <- pchisq(chi, df = xdf, lower.tail = FALSE)
  return(list("chi-squared" = chi, "degrees_of_freedom" = xdf, "p-value" = pv))
  }

```
